---
title: "Distance-based Redundancy Analysis"
author: "Katelynn"
date: "2024-05-29"
output: html_document
---

#### RDA Analysis on Drivers of Bloom Dynamics: 

Next for this analysis, I want to specify the drivers of cyanobacteria species...

- Select Cyanobacteria Species
- Select Environmental Variables
-- Wind Speed, Stability, Surface PAR, (DN, DP - eventually) TDN, DOC, Turbidity, Seston CN, Seston CP, MC_Q, ATX_Q

## Load Libraries
```{r}

library(ggforce)
library(vegan)
library(ggrepel)
library(BiodiversityR)
library(cimir)
library(ggpubr)
library(factoextra)

```

## Missisquoi Bay

```{r}

## selecting phytoplankton that are > 5%

MB_Phyto_Date <- Phytos %>%
  filter(Station == "MB") %>%
  select(SampleDate) %>%
  unique()
  
MB_Phyto_Keep <- Phytos %>%
  filter(Station == "MB") %>%
  group_by(Sample_ID, SampleDate, Genus) %>%
  summarize(RA_BM = sum(RA_BM)) %>%
  mutate(Keep = if_else(RA_BM >= 0.05, "Yes", "No")) %>%
  ungroup() %>%
  select(Genus, Keep) %>%
  filter(Keep=="Yes") %>%
  unique()

MB_Phyto_Matrix <- Phytos %>%
  filter(Taxonomic_group == "Cyanobacteria") %>%
  filter(Station == "MB") %>%
  filter(Genus %in% MB_Phyto_Keep$Genus) %>%
  select(SampleDate, Genus, BM_ug.L) %>%
  group_by(SampleDate, Genus) %>%
  summarise(BM = sum(BM_ug.L)) %>%
  pivot_wider(names_from = "Genus", values_from="BM") %>%
  filter(SampleDate > as.Date("2021-06-01")) %>%
  rename(Date = SampleDate)

MB_Phyto_Matrix[is.na(MB_Phyto_Matrix)] <- 0

MB_Seston_Matrix <- All_Combined %>%
  select(SestonP_uML, SestonC_uML, SestonN_uML, Station, Date) %>%
  filter(Station == "MB") %>%
  unique() %>%
  mutate(Seston_CP = SestonC_uML/SestonP_uML,
         Seston_CN = SestonC_uML/SestonN_uML, 
         Seston_NP = SestonN_uML/SestonP_uML) %>%
  na.omit() %>%
  select(-SestonP_uML, -SestonC_uML, -SestonN_uML, -Station) %>%
  mutate(Date = ymd(Date)) %>%
    filter(Date %in% MB_Phyto_Date$SampleDate)

## Weather Matrix

MB_Wind_RDA <- MB_Met %>%
  select(Wind_Speed, Date) %>%
  filter(Date %in% MB_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(Wind_Speed = mean(Wind_Speed))

MB_SS_RDA <- MB_Schmidt %>%
  mutate(Day = day(datetime),
         Month = month(datetime),
         Year = year(datetime),
         Date = make_date(Year, Month, Day)) %>%
  select(SS, Date) %>%
  filter(Date %in% MB_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(SS = mean(SS))


# Water Temp

MB_WT_RDA <- MB_Profiler %>%
  mutate(Day = day(datetime),
         Month = month(datetime),
         Year = year(datetime),
         Date = make_date(Year, Month, Day)) %>%
  select(Temp_C, Date) %>%
  filter(Date %in% MB_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(Temp_C = mean(Temp_C))

# Sum Stream Flow
MB_SF_RDA <- MB_SF_Stats %>%
  select(Sum_SF, Date) %>%
  filter(Date %in% MB_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(Sum_SF = mean(Sum_SF)) 

## Dissolved Nutrients. 

DissNut_S <- read_xlsx("./DissolvedNutrients_2021.xlsx") %>%
  filter(site == "Miss Bay") %>%
  select(site, depth, rep, date, `NO3_mg/L`, `NH4_mg/L`, `PO4_mg/L`) %>%
  mutate(Date = ymd(date),
        depth = if_else(depth == "surface", "Surface", depth),
        depth = if_else(depth == "bottom", "Bottom", depth)) %>%
  mutate(NO3 = if_else(`NO3_mg/L` < 0, 0, `NO3_mg/L`),
         NH4 = if_else(`NH4_mg/L`< 0, 0, `NH4_mg/L`),
         PO4 = if_else(`PO4_mg/L` < 0, 0, `PO4_mg/L`),
         NO3 = NO3 * 1000,
         NH4 = NH4 * 1000,
         PO4 = PO4 * 1000,
         DIN = NO3 + NH4) %>%
    na.omit() %>%
  group_by(site, depth, Date) %>%
  filter(depth=="Surface") %>%
  summarize(NH4 = mean(NH4),
            NO3 = mean(NO3),
            PO4_S = mean(PO4),
            DIN_S = mean(DIN)) %>%
  ungroup() %>%
  select(DIN_S, PO4_S, Date)

DissNut_B <- read_xlsx("./DissolvedNutrients_2021.xlsx") %>%
  filter(site == "Miss Bay") %>%
  select(site, depth, rep, date, `NO3_mg/L`, `NH4_mg/L`, `PO4_mg/L`) %>%
  mutate(Date = ymd(date),
        depth = if_else(depth == "surface", "Surface", depth),
        depth = if_else(depth == "bottom", "Bottom", depth)) %>%
  mutate(NO3 = if_else(`NO3_mg/L` < 0, 0, `NO3_mg/L`),
         NH4 = if_else(`NH4_mg/L`< 0, 0, `NH4_mg/L`),
         PO4 = if_else(`PO4_mg/L` < 0, 0, `PO4_mg/L`),
         NO3 = NO3 * 1000,
         NH4 = NH4 * 1000,
         PO4 = PO4 * 1000,
         DIN = NO3 + NH4) %>%
    na.omit() %>%
  group_by(site, depth, Date) %>%
  filter(depth=="Bottom") %>%
  summarize(NH4 = mean(NH4),
            NO3 = mean(NO3),
            PO4_B = mean(PO4),
            DIN_B = mean(DIN)) %>%
  ungroup() %>%
  select(DIN_B, PO4_B, Date)
## Join together. 
  
MB_Env_RDA <- full_join(MB_SF_RDA, MB_Wind_RDA) %>%
  full_join(., MB_WT_RDA) %>%
  full_join(., MB_SS_RDA) %>%
  full_join(., DissNut_S) %>%
  full_join(., DissNut_B) %>%
  #full_join(., MB_Seston_Matrix) %>%
  na.omit()


MB_Phyto_RDA <- MB_Phyto_Matrix %>%
  filter(Date %in% MB_Env_RDA$Date) %>%
  column_to_rownames(var="Date")

MB_Phyto_RDA[is.na(MB_Phyto_RDA)] <- 0


MB_Env_RDA <-   MB_Env_RDA %>%
  rownames_to_column(var="Remove") %>%
  select(-Remove) %>%
  column_to_rownames(var="Date")
```

## Checking for Colinearity in Variables: 

```{r}

library(corrplot)

MB_Corr = cor(MB_Env_RDA)
corrplot(MB_Corr, method = 'number')

corrplot::cor

```
Using a threshold of 0.75, we will remove Dissolved inorganic nitrogen in the bottom waters. Though the relationship between wind is really interestung. Suggesting internal N loading during low wind. 


## Run RDA. 

```{r}
library(BiodiversityR)

MB_Phyto_RDA <- MB_Phyto_RDA %>%
  select(-coccoid_cyano, -Gloeotrichia) %>%
  decostand("hellinger")

MB_Env_RDA <- MB_Env_RDA %>%
  decostand("standardize")


MB_dbrda <- dbrda(MB_Phyto_RDA ~ .,
                    data = MB_Env_RDA,
                  distance="bray")

print(MB_dbrda)

anova.cca(MB_dbrda)

```


Forward selection of the RDA


```{r}
# Initial RDA with ALL of the environmental data
MB_full <- dbrda(MB_Phyto_RDA ~ .,
                    data = MB_Env_RDA, 
                 distance="bray")

MB_null <- dbrda(MB_Phyto_RDA ~ 1, data=MB_Env_RDA, 
                 distance="bray")

modAIC <- MASS::stepAIC(MB_null,direction='forward', scope = list(lower = MB_null,
                                                              upper = MB_full))
summary(modAIC)


```
After running the forward selection model, bottom orthophosphate levels are the only important variable in Missisquoi Bay in Shaping the cyanobacteria community. 

## dbrda with selected variables
```{r}
MB_mod <- dbrda(MB_Phyto_RDA ~ PO4_B, data=MB_Env_RDA, 
                 distance="bray")

```

```{r}
library(BiodiversityR)

MB_mod_summary <- summary(MB_mod)

MB_mod_summary$concont
```


```{r}

## This code adds species scores back into the dbrda, since they are always missing. That way we can plot species on our ordination plot with the sites and environmental variables. 

sppscores(MB_mod) <- MB_Phyto_RDA

plot <- ordiplot(MB_mod, type = "point")

```

I don't really care about the specific samples here, I rather want to see the species and the environmental factors. Therefore, I am going to extract each and re-make this ordination plot in a ggplot environment.

```{r}
# This code chunk extracts data from the dbRDA

sites.long <- sites.long(plot, env.data=MB_Env_RDA) %>%
  select(PO4_B, axis1, axis2, labels)

axis.long <- axis.long(MB_mod, choices=c(1, 2))


df <- colSums(MB_Phyto_RDA) %>%
  as.data.frame() %>%
  mutate(`.` = abs(`.`))

species.long <- species.long(plot)
species.long

vectors.envfit <- envfit(plot, env=MB_Env_RDA)
vectors.long <- vectorfit.long(vectors.envfit) %>%
  filter(vector == "PO4_B")
vectors.long

```

Making the plot.

```{r}

MB_RDA <- ggplot() + 
      geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
      geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
      xlab(axis.long[1, "label"]) +
      ylab(axis.long[2, "label"]) +  
      scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
      scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +    
      geom_point(data=sites.long, 
                 aes(x=axis1, y=axis2), 
                 size=1) +
      #geom_point(data=species.long_filt, 
      #          aes(x=axis1, y=axis2), size=1.5, color="dodgerblue3", alpha=0.6) +
      geom_text_repel(data=species.long, 
                    aes(x=axis1*1, y=axis2*1, label=labels),
                    color="dodgerblue4", size=2) +
      geom_segment(data=vectors.long,
                 aes(x=0, y=0, xend=axis1*1, yend=axis2*1), 
                 colour="black", size=0.5, arrow=arrow(type="open", length=unit(0.15, "cm"))) +
      geom_text_repel(data=vectors.long, 
                    aes(x=axis1*1, y=axis2*1, label=vector),
                    size=2,
                    colour="black") +
      ggsci::scale_colour_npg() +
      theme_classic() +
      coord_fixed(ratio=1) +
  ggtitle("A")


```

### St. Albans Bay. 
```{r}

## selecting phytoplankton that are > 5%

SA_Phyto_Date <- Phytos %>%
  filter(Station == "SA") %>%
  select(SampleDate) %>%
  unique()
  
SA_Phyto_Keep <- Phytos %>%
  filter(Station == "SA") %>%
  group_by(Sample_ID, SampleDate, Genus) %>%
  summarize(RA_BM = sum(RA_BM)) %>%
  mutate(Keep = if_else(RA_BM >= 0.05, "Yes", "No")) %>%
  ungroup() %>%
  select(Genus, Keep) %>%
  filter(Keep=="Yes") %>%
  unique()

SA_Phyto_Matrix <- Phytos %>%
  filter(Taxonomic_group == "Cyanobacteria") %>%
  filter(Station == "SA") %>%
  filter(Genus %in% SA_Phyto_Keep$Genus) %>%
  select(SampleDate, Genus, BM_ug.L) %>%
  group_by(SampleDate, Genus) %>%
  summarise(BM = sum(BM_ug.L)) %>%
  pivot_wider(names_from = "Genus", values_from="BM") %>%
  filter(SampleDate > as.Date("2021-06-01")) %>%
  rename(Date = SampleDate)

SA_Phyto_Matrix[is.na(SA_Phyto_Matrix)] <- 0

SA_Seston_Matrix <- All_Combined %>%
  select(SestonP_uML, SestonC_uML, SestonN_uML, Station, Date) %>%
  filter(Station == "SA") %>%
  unique() %>%
  mutate(Seston_CP = SestonC_uML/SestonP_uML,
         Seston_CN = SestonC_uML/SestonN_uML, 
         Seston_NP = SestonN_uML/SestonP_uML) %>%
  na.omit() %>%
  select(-SestonP_uML, -SestonC_uML, -SestonN_uML, -Station) %>%
  mutate(Date = ymd(Date)) %>%
    filter(Date %in% SA_Phyto_Date$SampleDate)

## Weather Matrix

SA_Wind_RDA <- SA_Met %>%
  mutate(Day = day(Datetime)) %>%
  mutate(Date = make_date(Year, Month, Day)) %>%
  select(Wind_Speed_m.s, Date) %>%
  filter(Date %in% SA_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(Wind_Speed = mean(Wind_Speed_m.s))

SA_SS_RDA <- SA_Schmidt %>%
  mutate(Day = day(datetime),
         Month = month(datetime),
         Year = year(datetime),
         Date = make_date(Year, Month, Day)) %>%
  select(SS, Date) %>%
  filter(Date %in% SA_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(SS = mean(SS))


# Water Temp

SA_WT_RDA <- SA_Profiler %>%
  mutate(Day = day(datetime),
         Month = month(datetime),
         Year = year(datetime),
         Date = make_date(Year, Month, Day)) %>%
  select(Temp_C, Date) %>%
  filter(Date %in% SA_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(Temp_C = mean(Temp_C))

# Sum Stream Flow
SA_SF_RDA <- SA_SF_Stats %>%
  select(Sum_SF, Date) %>%
  filter(Date %in% SA_Phyto_Date$SampleDate) %>%
  group_by(Date) %>%
  summarize(Sum_SF = mean(Sum_SF)) 

## Dissolved Nutrients. 

DissNut_S <- read_xlsx("./DissolvedNutrients_2021.xlsx") %>%
  filter(site != "Miss Bay") %>%
  select(site, depth, rep, date, `NO3_mg/L`, `NH4_mg/L`, `PO4_mg/L`) %>%
  mutate(Date = ymd(date),
        depth = if_else(depth == "surface", "Surface", depth),
        depth = if_else(depth == "bottom", "Bottom", depth)) %>%
  mutate(NO3 = if_else(`NO3_mg/L` < 0, 0, `NO3_mg/L`),
         NH4 = if_else(`NH4_mg/L`< 0, 0, `NH4_mg/L`),
         PO4 = if_else(`PO4_mg/L` < 0, 0, `PO4_mg/L`),
         NO3 = NO3 * 1000,
         NH4 = NH4 * 1000,
         PO4 = PO4 * 1000,
         DIN = NO3 + NH4) %>%
    na.omit() %>%
  group_by(site, depth, Date) %>%
  filter(depth=="Surface") %>%
  summarize(NH4 = mean(NH4),
            NO3 = mean(NO3),
            PO4_S = mean(PO4),
            DIN_S = mean(DIN)) %>%
  ungroup() %>%
  select(DIN_S, PO4_S, Date)

DissNut_B <- read_xlsx("./DissolvedNutrients_2021.xlsx") %>%
  filter(site != "Miss Bay") %>%
  select(site, depth, rep, date, `NO3_mg/L`, `NH4_mg/L`, `PO4_mg/L`) %>%
  mutate(Date = ymd(date),
        depth = if_else(depth == "surface", "Surface", depth),
        depth = if_else(depth == "bottom", "Bottom", depth)) %>%
  mutate(NO3 = if_else(`NO3_mg/L` < 0, 0, `NO3_mg/L`),
         NH4 = if_else(`NH4_mg/L`< 0, 0, `NH4_mg/L`),
         PO4 = if_else(`PO4_mg/L` < 0, 0, `PO4_mg/L`),
         NO3 = NO3 * 1000,
         NH4 = NH4 * 1000,
         PO4 = PO4 * 1000,
         DIN = NO3 + NH4) %>%
    na.omit() %>%
  group_by(site, depth, Date) %>%
  filter(depth=="Bottom") %>%
  summarize(NH4 = mean(NH4),
            NO3 = mean(NO3),
            PO4_B = mean(PO4),
            DIN_B = mean(DIN)) %>%
  ungroup() %>%
  select(DIN_B, PO4_B, Date)
## Join together. 
  
SA_Env_RDA <- full_join(SA_SF_RDA, SA_Wind_RDA) %>%
  full_join(., SA_WT_RDA) %>%
  full_join(., SA_SS_RDA) %>%
  full_join(., DissNut_S) %>%
  full_join(., DissNut_B) %>%
  #full_join(., MB_Seston_Matrix) %>%
  na.omit()


SA_Phyto_RDA <- SA_Phyto_Matrix %>%
  filter(Date %in% SA_Env_RDA$Date) %>%
  column_to_rownames(var="Date")

SA_Phyto_RDA[is.na(SA_Phyto_RDA)] <- 0


SA_Env_RDA <-   SA_Env_RDA %>%
  rownames_to_column(var="Remove") %>%
  select(-Remove) %>%
  column_to_rownames(var="Date")
```

## Checking for Colinearity in Variables: 

```{r}

library(corrplot)

SA_Corr = cor(SA_Env_RDA)
corrplot(SA_Corr, method = 'number')

corrplot::cor

```


## Run RDA. 

```{r}
library(BiodiversityR)

SA_Phyto_RDA <- SA_Phyto_RDA %>%
  decostand("hellinger")

SA_Env_RDA <- SA_Env_RDA %>%
  decostand("standardize")


SA_dbrda <- dbrda(SA_Phyto_RDA ~ .,
                    data = SA_Env_RDA,
                  distance="bray")

print(SA_dbrda)

anova.cca(SA_dbrda)

```


Forward selection of the RDA


```{r}
# Initial RDA with ALL of the environmental data
SA_full <- dbrda(SA_Phyto_RDA ~ .,
                    data = SA_Env_RDA, 
                 distance="bray")

SA_null <- dbrda(SA_Phyto_RDA ~ 1, data=SA_Env_RDA, 
                 distance="bray")

modAIC <- MASS::stepAIC(SA_null,direction='forward', scope = list(lower = SA_null,
                                                              upper = SA_full))


SA_mod <- dbrda(SA_Phyto_RDA ~ DIN_S + DIN_B + Wind_Speed + Temp_C + PO4_S + 
    SS + Sum_SF, 
                data=SA_Env_RDA, distance="bray")

print(SA_mod)

anova.cca(SA_mod)

```
After running the forward selection model, bottom orthophosphate levels are the only important variable in Missisquoi Bay in Shaping the cyanobacteria community. 

```{r}
library(BiodiversityR)

SA_mod_summary <- summary(SA_mod)

SA_mod_summary$concont

```


```{r}

## This code adds species scores back into the dbrda, since they are always missing. That way we can plot species on our ordination plot with the sites and environmental variables. 

sppscores(SA_mod) <- SA_Phyto_RDA

plot <- ordiplot(SA_mod, type = "point")

```


```{r}
# This code chunk extracts data from the dbRDA

sites.long <- sites.long(plot, env.data=SA_Env_RDA)

axis.long <- axis.long(SA_mod, choices=c(1, 2))


df <- colSums(SA_Phyto_RDA) %>%
  as.data.frame() %>%
  mutate(`.` = abs(`.`))


species.long <- species.long(plot)
species.long


vectors.envfit <- envfit(plot, env=SA_Env_RDA)
vectors.long <- vectorfit.long(vectors.envfit)
vectors.long

## Stream Flow is not significant. Everything else is. 

```

Making the plot.

```{r}

library(ggforce)
library(vegan)
library(ggrepel)
library(BiodiversityR)
library(cimir)
library(ggpubr)
library(factoextra)


SA_RDA <- ggplot() + 
      geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
      geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
      xlab(axis.long[1, "label"]) +
      ylab(axis.long[2, "label"]) +  
      scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
      scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +    
      geom_point(data=sites.long, 
                 aes(x=axis1, y=axis2), 
                 size=1) +
      #geom_point(data=species.long_filt, 
      #          aes(x=axis1, y=axis2), size=1.5, color="dodgerblue3", alpha=0.6) +
      geom_text_repel(data=species.long, 
                    aes(x=axis1*1, y=axis2*1, label=labels),
                    color="dodgerblue4", size=2) +
      geom_segment(data=vectors.long,
                 aes(x=0, y=0, xend=axis1*1, yend=axis2*1), 
                 colour="black", size=0.5, arrow=arrow(type="open", length=unit(0.15, "cm"))) +
      geom_text_repel(data=vectors.long, 
                    aes(x=axis1*1, y=axis2*1, label=vector),
                    size=2,
                    colour="black") +
      ggsci::scale_colour_npg() +
      theme_classic() +
      coord_fixed(ratio=1) +
  ggtitle("B")


```
